---
title: "Rosiglitazone effectiveness"
author: "Rosauro and Joe"
date: "April 10, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, 
               echo = FALSE, 
               warning = FALSE, 
               message = FALSE, 
               error = TRUE, 
               cache = FALSE)
options(scipen = '999')
```

```{r}
if(!require('tidyverse')){
  install.packages('tidyverse')
}
if(!require('ggplot2')){
  install.packages('ggplot2')
}
if(!require('readxl')){
  install.packages('readxl')
}
if(!require('readxl')){
  install.packages('readxl')
}
if(!require('RColorBrewer')){
  install.packages('RColorBrewer')
}
if(!require('plotly')){
  install.packages('plotly')
}
if(!require('ggthemes')){
  install.packages('ggthemes')
}
```

# Visualizations

```{r}
# Read data into R
df <- read_excel('Joe-Rosi.xlsx',
                 skip = 0)

# Make long
df <- gather(df, time, value, `0 h`:`72h`)

# Get hours
df$hours <- as.numeric(gsub('h| h', '', df$time))

# MAke value numeric
df$value <- as.numeric(df$value)

# Get value as percent of first
df <- df %>%
  group_by(`Study number`) %>%
  mutate(p = value / first(value) * 100) %>%
  ungroup

# Arrange by hours
df <- df %>%
  arrange(hours)

# # manually set to 0 parisetemia
# df$value[df$key == 'Parasitemia' &
#            df$hours == 216] <- 0
# df$p[df$key == 'Parasitemia' &
#            df$hours == 216] <- 0

# # Remove NA values
# df <- df %>%
#   filter(!is.na(p))


# # Filter down
# df <- df %>%
#   filter(key %in% c('Hemoglobin', 'Parasitemia', 'White Blood Cells'))
# 
# 
# # Multiply WBC by 1000
# df$value[df$key == 'White Blood Cells'] <-
#   df$value[df$key == 'White Blood Cells'] * 1000
# 
# # plot
# cols <- colorRampPalette(brewer.pal(n = 9, name = 'Set1'))(length(unique(df$key)))
# 
# # Make a sub dataframe for labeling hgb cut points
# hgb <- df %>% filter(grepl('Hemoglobin', key))
# hgb$time <- as.numeric(gsub('h', '', hgb$time))
# # need to interpolate the value at the unknown hours
# left <- data_frame(time = seq(0, max(hgb$time), by = 1))
# hgb <- left_join(left, hgb)
# hgb$p <- zoo::na.approx(object = hgb$p,
#                             x = hgb$time)
# hgb$value <- zoo::na.approx(object = hgb$value,
#                         x = hgb$time)
# hgb <- hgb %>% filter(time %in% c(6, 98, 174))
# hgb$hours <- hgb$time
# hgb$time <- paste0(hgb$time, 'h')
# 
# # Get better labels
# 
# greeks <- list(
#   bquote('HGB (g/dL)'),
#   bquote(.('Parasitemia (parasites /') ~ mu*.('L)')),
#   # bquote(.('joe'),mu),
#   bquote(.('WBCC (Leococytes /') ~ mu*.('L)')))
# 
# greeks2 <- 
#   c('HGB (g/dL)',
#     'Parasitemia (parasites/microlitre)',
#     'WBCC (Leococytes/microlitre')
# 
# # plot
# cols <- colorRampPalette(brewer.pal(n = 9, name = 'Set1'))(length(unique(df$key)))
# 
# # Add the points
# hgb$key <- 'Hemoglobin'
# hgb$triangle <- hgb$value
# 
# df <- bind_rows(hgb, df)

# Perform interpolation
study_numbers <- sort(unique(df$`Study number`))
left <- expand.grid(hours = seq(0, max(df$hours, na.rm = TRUE), by = 1),
                   `Study number` = study_numbers)
right <- df %>%
  dplyr::select(`Study number`, Treatment)

joined <- left_join(left, df %>%
                      dplyr::select(`Study number`,
                                    hours,
                                    value, 
                                    p))
joined <- left_join(joined,
                    right)
# for (i in 1:length(study_numbers)){
#   this_study_number <- study_numbers[i]
#   these_data <- df %>%
#     filter(`Study number` == this_study_number)
#   these_data$p <- zoo::na.approx(object = these_data$p,
#                                  x = these_data$hours,
#                                  na.rm = FALSE)
# }

df <- joined %>%
  group_by(`Study number`) %>%
  mutate(value = zoo::na.approx(object = value,
                                x = hours,
                                na.rm = FALSE),
         p = zoo::na.approx(object = p,
                                x = hours,
                            na.rm = FALSE))

# Group by treatment
bt <- df %>%
  group_by(Treatment, hours) %>%
  summarise(q25 = quantile(value, 0.25, na.rm = TRUE),
            q50 = quantile(value, 0.5, na.rm = TRUE),
            q50p = quantile(p, 0.5, na.rm = TRUE),
            q75 = quantile(value, 0.75, na.rm = TRUE),
            value = mean(value,
                         na.rm = TRUE),
            p = mean(p, na.rm = TRUE),
            avg_p = mean(p, na.rm = TRUE),
            avg = mean(value, na.rm = TRUE))


```

The below chart shows the _median_ parasitemia (as percentage of value at hour 0) by treatment.

```{r}
ggplot(data = bt,
       aes(x = hours,
           y = q50p,
           group = Treatment,
           color = Treatment)) +
  geom_line() +
  theme_hc() +
  labs(x = 'Hours after recruitment',
       y = 'Parasitemia percentage of hour 0',
       title = 'Parasitemia: median: Placebo vs. Rosiglitazone') +
  geom_vline(xintercept = seq(0, 100, 10),
             lty = 2,
             alpha =0.2)
```

The below chart is identical to the above, but shows true parasitemia values (rather than a percentage of hour 0).

```{r}
ggplot(data = bt,
       aes(x = hours,
           y = q50,
           group = Treatment,
           color = Treatment)) +
  geom_line() +
  theme_hc() +
  labs(x = 'Hours after recruitment',
       y = 'Parasitemia',
       title = 'Parasitemia: median: Placebo vs. Rosiglitazone') +
  geom_vline(xintercept = seq(0, 100, 10),
             lty = 2,
             alpha =0.2)
```


As an alternative to the median (which can mask magnitude in extremes), we can also explore the _mean_ (below two charts).


```{r}
ggplot(data = bt,
       aes(x = hours,
           y = avg_p,
           group = Treatment,
           color = Treatment)) +
  geom_line() +
  theme_hc() +
  labs(x = 'Hours after recruitment',
       y = 'Parasitemia percentage of hour 0',
       title = 'Parasitemia: mean: Placebo vs. Rosiglitazone') +
  geom_vline(xintercept = seq(0, 100, 10),
             lty = 2,
             alpha =0.2)
```


```{r}
ggplot(data = bt,
       aes(x = hours,
           y = avg,
           group = Treatment,
           color = Treatment)) +
  geom_line() +
  theme_hc() +
  labs(x = 'Hours after recruitment',
       y = 'Parasitemia',
       title = 'Parasitemia: mean: Placebo vs. Rosiglitazone') +
  geom_vline(xintercept = seq(0, 100, 10),
             lty = 2,
             alpha =0.2)
```


All summary statistics (median, mean, etc.) can mask important individual variation. The below chart shows one line for each individual, colored by treatment protocol.

```{r}
ggplot(data = df,
       aes(x = hours,
           y = value,
           color = Treatment,
           group = `Study number`)) +
  geom_line(alpha = 0.6) +
  theme_hc() +
  labs(x = 'Hours',
       y = 'Parasitemia',
       title = 'Improvement trajectories')
```

The below chart is identical to the above, but the y-axis is a percentage of value at hour 0.

```{r}
ggplot(data = df,
       aes(x = hours,
           y = p,
           color = Treatment,
           group = `Study number`)) +
  geom_line(alpha = 0.6) +
  theme_hc() +
  labs(x = 'Hours',
       y = 'Parasitemia: percentage of hour 0',
       title = 'Improvement trajectories') 

```

# Hypothesis test

For now, we'll use a simple linear model to estimate the effect (and significance) of treatment on parasitemia, adjusted for time elapsed and parasitemia at hour 0.

```{r}
model_data <- df %>%
  arrange(hours) %>%
  group_by(id = `Study number`) %>%
  mutate(start_value = first(value)) %>%
  ungroup

fit <- lm(value ~ Treatment + start_value,
          data = model_data)

out <- broom::tidy(fit)

out$estimate <- round(out$estimate, digits = 3)
out$std.error <- round(out$std.error, digits = 3)
out$statistic <- round(out$statistic, digits = 3)
out$p.value <- round(out$p.value, digits = 6)
DT::datatable(out)
```

This simple approach is subject to a great deal of bias (non-linearity, outliers), but gives us an initial (and reproducible) method for assessing significance.